Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?

### 2018-09-18 analysis/paper/paper.Rmd
Update pkgdown site
Fixes analysis document
Fixes captions in the analysis document and uses new figures from unthinned MCMC

---
title: Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au]  \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjorn Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute, Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland, Centre for Crop Health, West Street, Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two management practices may trigger sheath blight of rice, caused by _Rhizoctonia solani_, which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split-plot experiments with a water saving (alternate wetting and drying) regime and traditional flood irrigation regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored in the same way in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did differ significantly by irrigation treatment, but leaf sheath blight severity did not. Our findings suggest that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross-cutting research in this area is warranted.
bibliography: references.bib
output:
  bookdown::word_document2:
    toc: no
    fig_caption: yes
    reference_docx: "../templates/template.docx"
  always_allow_html: yes
header-includes:
   \usepackage{endfloat}
   \usepackage{setspace}\doublespacing
   \usepackage{lineno}
   \linenumbers
---

@@

library(bookdown)
library(captioner)
library(knitr)
library(ggplot2)
library(ggpubr)
library(grid)
library(gridExtra)
library(rice.awd.pests)
```

@@

The split-plot treatment rates of N were N0 (no N supply), N100 (100 kg per ha applied as urea in three splits at final harrowing, active tillering and panicle initiation) and N120 (120 kg per ha applied as urea in three splits at final harrowing, active tillering and panicle initiation) (Tab. \@ref(tab:tab_1)).

@@

Trays of NSIC Rc222 were seeded on 19 December 2014. NSIC Rc222 is an inbred variety released by the Philippine Rice Research Institute (PhilRice), with a 114 day maturity when transplanted were established. The variety is commonly grown by farmers in the area, having moderate resistance to brown plant hopper, green leaf hopper and yellow stem borer, but susceptibility to tungro. Plots were established by manually transplanting seedlings on 9 January 2015 in hills with six to eight seedlings per hill and a distance of 20cm within and between rows.

@@

Split-plot N rates differed from the 2015 rates with only two N treatments, N60 (60 kg per ha as urea split into two applications) and N180 (180 kg per ha in three splits) being applied (Tab. \@ref(tab:tab_1)).

@@

Four MCMC chains were run for 55,000 steps with the first 5000 discarded as burn in. The remaining 50,000 samples were thinned by taking every tenth sample resulting in 5,000 independent draws from the posterior distribution of the parameters of the model. The `coda` [@coda2006] package was used to provide utilities to check MCMC chain convergence using visual inspection of resulting trace graphs and the Gelman-Rubin test. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://github.com/openplantpathology/rice_awd_pests> [@Sparks2018].

@@

In 2015 the incidence of tiller ShB remained low throughout the growing season (Fig. \@ref(fig:fig_1)a, \@ref(fig:fig_2)a). Water management was not significantly different (Fig. \@ref(fig:fig_3)a, \@ref(fig:fig_5)a). However, the N treatments, N100 and N120, were both significantly different when compared with the control N0 treatment (Fig. \@ref(fig:fig_4)a, \@ref(fig:fig_5)a).

@@

Tiller ShB severity remained below 2% (Fig. \@ref(fig:fig_2)c, \@ref(fig:fig_1)c). Both the N100 and N120 treatments were significantly different than the control N0 treatment (Fig. \@ref(fig:fig_4)c, \@ref(fig:fig_5)c). However, the AWD water management was not significantly different from the FLD treatment (Fig. \@ref(fig:fig_3)c, \@ref(fig:fig_5)c).

@@

Severity of leaf ShB remained low, less than 0.4% across all treatments (Fig. \@ref(fig:fig_1)e, \@ref(fig:fig_2)e). Both the N100 and N120 treatments were significantly different than the N0 treatment (Fig. \@ref(fig:fig_4)e, \@ref(fig:fig_5)e). However, the AWD water management was not significantly different from the FLD treatment (Fig. \@ref(fig:fig_3)e, \@ref(fig:fig_5)e).

@@

The changes to the inoculation methodology resulted in a higher rate of infection in 2016, with the N180 treatment reaching a maximum value of 98% incidence at the third observation (Fig. \@ref(fig:fig_2)b). The N treatment N180, was significantly different than the control N60 treatment (Fig. \@ref(fig:fig_4)b, \@ref(fig:fig_5)b). As in the 2015 study, water management did not significantly differ (Fig. \@ref(fig:fig_3)b, \@ref(fig:fig_5)b).

@@

As with the tiller incidence, the tiller severity increased with the changed inoculation methods with a maximum of 7.6% for the flooded treatment (Fig. \@ref(fig:fig_2)d). 
In 2016 the N180 treatment was significantly different from the N60 treatment, with N180 severity being higher (Fig. \@ref(fig:fig_4)d, \@ref(fig:fig_5)d). The AWD water management, which was lower than the FLD treatment, was also significantly different (Fig. \@ref(fig:fig_3)d, \@ref(fig:fig_5)d).

@@

In 2016 the neither of the treatments, N rate or water management, were significantly different from the control treatments for leaf ShB severity (Fig. \@ref(fig:fig_3)f, \@ref(fig:fig_5)f, \@ref(fig:fig_4)f, \@ref(fig:fig_5)f).

@@

In both experiments we were unable to detect any significant effect of AWD on ShB that would cause increases in the disease, which could hinder adoption of the technology. In only one instance were we able to detect any effect of AWD on ShB. In the 2016 experiment the tiller ShB severity was significantly lower for AWD than FLD (Fig. \@ref(fig:fig_3)d, \@ref(fig:fig_5)d), indicating a possible adverse effect of using AWD on tiller ShB severity under conditions where there is high sheath blight pressure.

@@

The changes made for the 2016 experiment appear to have improved the experiment, increasing the incidence and decreased variability of sheath blight in the plots (Fig. \@ref(fig:fig_1)a:b, \@ref(fig:fig_2)c:d, \@ref(fig:fig_3)a:b, \@ref(fig:fig_4)c:d). The rice establishment method is known to affect the spread of sheath blight [@Willocquet2000]. It is possible that the mechanical transplanter could increase ShB spread in the canopy as compared with traditional manual transplanting. This however was not investigated as a part of this study but could bear further research.

@@

```{r table-caps}
table_nums <- captioner(prefix = "Tab.")

tab.1_cap <- table_nums(name = "tab_1", 
                        caption = "Nitrogen application rates for 2015 and 2016. Five different nitrogen treatments were applied depending on the experiment. In the 2015 experiment three nitrogen rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two nitrogen rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180). Treatments were applied in splits, basal, tillering and panicle initiation growth stages.")
```

@@

`r table_nums("tab_1")`
```{r N-rates, fig.cap=tab.1_cap}
kable(N_rates)

@@

```{r figure-caps}
figure_nums <- captioner(prefix = "Fig.")

fig.1_cap <- figure_nums(name = "fig_1", 
                        caption = "Sheath blight progress for 2015 and 2016 experiments. Disease progress was measured in the 2015 experiment at five assessment points and 2016 experiment at four assessment points during the dry season both years. Main plot irrigation treatments were permanently flooded (FLD) and alternate wetting and drying (AWD). Points represent the mean of four replications.")

fig.2_cap <- figure_nums(name = "fig_2", 
                        caption = "Sheath blight incidence progress was measured in the 2015 experiment at five assessment points and 2016 experiment at four assessment points during the dry season both years. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180). Points represent the mean of four replications.")

fig.3_cap <- figure_nums(name = "fig_3", 
                        caption = "Sheath blight progress was measured in the 2015 at five assessment points and 2016 experiment at four points. Sheath blight incidence was rated as the number of infected tillers divided by the total number of tillers per hill and used to calculate the area under the disease progress stairs (AUDPS). Sheath blight severity was rated on an unevenly spaced categorical scale and converted to the mid-point percentage value to calculate the AUDPS. Main plot irrigation treatments were permanently flooded (FLD) and alternate wetting and drying (AWD).")

fig.4_cap <- figure_nums(name = "fig_4", 
                        caption = "Sheath blight severity progress was measured in the 2015 at five assessment points and 2016 experiment at four points. Sheath blight incidence was rated as the number of infected tillers divided by the total number of tillers per hill and used to calculate the area under the disease progress stairs (AUDPS). Sheath blight severity was rated on an unevenly spaced categorical scale and converted to the mid-point percentage value to calculate the AUDPS. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180).")

fig.5_cap <- figure_nums(name = "fig_5", 
                        caption = "Posterior means and 95% credible intervals for the explanatory variables used in models of alternate wetting and drying irrigations (AWD) and N rate (NRTE) on sheath blight tiller incidence and tiller severity. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180).")

fig.6_cap <- figure_nums(name = "fig_6", 
                        caption = "Posterior means and 95% credible intervals for the explanatory variables used in models of alternate wetting and drying irrigations (AWD) and N rate (NRTE) and the interaction of the two treatments on sheath blight leaf severity. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180).")
```

`r figure_nums("fig_1")`
```{r fig_1, fig.cap="fig.1_cap"}

@@

##### ShB N progress
`r figure_nums("fig_2")`
```{r fig_2, fig.cap="fig.2_cap"}

@@

`r figure_nums("fig_3")`
```{r fig_3, fig.cap="fig.3_cap"}

@@

##### ShB boxplot - NRTE
`r figure_nums("fig_4")`
```{r fig_4, fig.cap="fig.4_cap"}

@@

`r figure_nums("fig_5")`
```{r fig_5, fig.cap="fig.5_cap"}

@@

`r figure_nums("fig_6")`
```{r fig_6, fig.cap="fig.5_cap"}

>>>>

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au]  \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjorn Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute, Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland, Centre for Crop Health, West Street, Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two management practices may trigger sheath blight of rice, caused by _Rhizoctonia solani_, which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split-plot experiments with a water saving (alternate wetting and drying) regime and traditional flood irrigation regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored in the same way in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did differ significantly by irrigation treatment, but leaf sheath blight severity did not. Our findings suggest that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross-cutting research in this area is warranted.
bibliography: references.bib
output:
  bookdown::word_document2:
    toc: no
    reference_docx: "../templates/template.docx"
---

@@

library(bookdown)
library(knitr)
library(ggplot2)
library(ggpubr)
library(grid)
library(gridExtra)
library(rice.awd.pests)
```

@@

The split-plot treatment rates of N were N0 (no N supply), N100 (100 kg per ha applied as urea in three splits at final harrowing, active tillering and panicle initiation) and N120 (120 kg per ha applied as urea in three splits at final harrowing, active tillering and panicle initiation) (Tab. \@ref(tab:tab-1)).

@@

Trays of NSIC Rc222 were seeded on 19 December 2014. NSIC Rc222 is an inbred variety released by the Philippine Rice Research Institute (PhilRice), with a 114 day maturity when transplanted, were established. The variety is commonly grown by farmers in the area, having moderate resistance to brown plant hopper, green leaf hopper and yellow stem borer, but susceptibility to tungro. Plots were established by manually transplanting seedlings on 9 January 2015 in hills with six to eight seedlings per hill and a distance of 20cm within and between rows.

@@

Split-plot N rates differed from the 2015 rates with only two N treatments, N60 (60 kg per ha as urea split into two applications) and N180 (180 kg per ha in three splits) being applied (Tab. \@ref(tab:tab-1)).

@@

Four MCMC chains were run for 55,000 steps with the first 5000 discarded as burn in. The remaining 50,000 samples were used to determine the posterior distribution of the parameters of the model. The `coda` [@coda2006] package was used to provide utilities to check MCMC chain convergence using visual inspection of resulting trace graphs and the Gelman-Rubin test. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://github.com/openplantpathology/rice_awd_pests> [@Sparks2018].

@@

In 2015 the incidence of tiller ShB remained low throughout the growing season (Fig. \@ref(fig:fig-1)a, \@ref(fig:fig-2)a). Water management was not significantly different (Fig. \@ref(fig:fig-3)a, \@ref(fig:fig-5)a). However, the N treatments, N100 and N120, were both significantly different when compared with the control N0 treatment (Fig. \@ref(fig:fig-4)a, \@ref(fig:fig-5)a).

@@

Tiller ShB severity remained below 2% (Fig. \@ref(fig:fig-2)c, \@ref(fig:fig-1)c). Both the N100 and N120 treatments were significantly different than the control N0 treatment (Fig. \@ref(fig:fig-4)c, \@ref(fig:fig-5)c). However, the AWD water management was not significantly different from the FLD treatment (Fig. \@ref(fig:fig-3)c, \@ref(fig:fig-5)c).

@@

Severity of leaf ShB remained low, less than 0.4% across all treatments (Fig. \@ref(fig:fig-1)e, \@ref(fig:fig-2)e). Both the N100 and N120 treatments were significantly different than the N0 treatment (Fig. \@ref(fig:fig-4)e, \@ref(fig:fig-5)e). However, the AWD water management was not significantly different from the FLD treatment (Fig. \@ref(fig:fig-3)e, \@ref(fig:fig-5)e).

@@

The changes to the inoculation methodology resulted in a higher rate of infection in 2016, with the N180 treatment reaching a maximum value of 98% incidence at the third observation (Fig. \@ref(fig:fig-2)b). The N treatment N180, was significantly different than the control N60 treatment (Fig. \@ref(fig:fig-4)b, \@ref(fig:fig-5)b). As in the 2015 study, water management did not significantly differ (Fig. \@ref(fig:fig-3)b, \@ref(fig:fig-5)b).

@@

As with the tiller incidence, the tiller severity increased with the changed inoculation methods with a maximum of 7.6% for the flooded treatment (Fig. \@ref(fig:fig-2)d). 
In 2016 the N180 treatment was significantly different from the N60 treatment, with N180 severity being higher (Fig. \@ref(fig:fig-4)d, \@ref(fig:fig-5)d). The AWD water management, which was lower than the FLD treatment, was also significantly different (Fig. \@ref(fig:fig-3)d, \@ref(fig:fig-5)d).

@@

In 2016 the neither of the treatments, N rate or water management, were significantly different from the control treatments for leaf ShB severity (Fig. \@ref(fig:fig-3)f, \@ref(fig:fig-5)f, \@ref(fig:fig-4)f, \@ref(fig:fig-5)f).

@@

In both experiments we were unable to detect any significant effect of AWD on ShB that would cause increases in the disease, which could hinder adoption of the technology. In only one instance were we able to detect any effect of AWD on ShB. In the 2016 experiment the tiller ShB severity was significantly lower for AWD than FLD (Fig. \@ref(fig:fig-3)d, \@ref(fig:fig-5)d), indicating a possible adverse effect of using AWD on tiller ShB severity under conditions where there is high sheath blight pressure.

@@

The changes made for the 2016 experiment appear to have improved the experiment, increasing the incidence and decreased variability of sheath blight in the plots (Fig. \@ref(fig:fig-1)a:b, \@ref(fig:fig-2)c:d, \@ref(fig:fig-3)a:b, \@ref(fig:fig-4)c:d). The rice establishment method is known to affect the spread of sheath blight [@Willocquet2000]. It is possible that the mechanical transplanter could increase ShB spread in the canopy as compared with traditional manual transplanting. This however was not investigated as a part of this study but could bear further research.

@@

@@

```{r N-rates}

kable(N_rates,
      caption = "Nitrogen application rates for 2015 and 2016. Five different nitrogen treatments were applied depending on the experiment. In the 2015 experiment three nitrogen rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two nitrogen rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180). Treatments were applied in splits, basal, tillering and panicle initiation growth stages.")

@@

```{r fig-1, fig.cap="Sheath blight progress for 2015 and 2016 experiments. Disease progress was measured in the 2015 experiment at five assessment points and 2016 experiment at four assessment points during the dry season both years. Main plot irrigation treatments were permanently flooded (FLD) and alternate wetting and drying (AWD). Points represent the mean of four replications."}

@@

##### ShB N progress

```{r fig-2, fig.cap="Sheath blight incidence progress was measured in the 2015 experiment at five assessment points and 2016 experiment at four assessment points during the dry season both years. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180). Points represent the mean of four replications."}

@@

```{r fig-3, fig.cap="Sheath blight progress was measured in the 2015 at five assessment points and 2016 experiment at four points. Sheath blight incidence was rated as the number of infected tillers divided by the total number of tillers per hill and used to calculate the area under the disease progress stairs (AUDPS). Sheath blight severity was rated on an unevenly spaced categorical scale and converted to the mid-point percentage value to calculate the AUDPS. Main plot irrigation treatments were permanently flooded (FLD) and alternate wetting and drying (AWD)."}

@@

##### ShB boxplot - NRTE
```{r fig-4, fig.cap="Sheath blight severity progress was measured in the 2015 at five assessment points and 2016 experiment at four points. Sheath blight incidence was rated as the number of infected tillers divided by the total number of tillers per hill and used to calculate the area under the disease progress stairs (AUDPS). Sheath blight severity was rated on an unevenly spaced categorical scale and converted to the mid-point percentage value to calculate the AUDPS. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180)."}

@@

```{r fig-5, fig.cap="Posterior means and 95% credible intervals for the explanatory variables used in models of alternate wetting and drying irrigations (AWD) and N rate (NRTE) on sheath blight tiller incidence and tiller severity. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180)."}

@@

```{r fig-6, fig.cap="Posterior means and 95% credible intervals for the explanatory variables used in models of alternate wetting and drying irrigations (AWD) and N rate (NRTE) and the interaction of the two treatments on sheath blight leaf severity. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180)."}
